A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-02 tex/codatadp.tex
conventions

\newcommand\Data{\Types_{\mathrm{dt}}}
\newcommand\Codata{\Types_{\mathrm{codt}}}
\newcommand\Nondata{\Types_{\mathrm{other}}}

@@

  * less briefly: codatatypes and why they are useful
    * in Agda, Coq, Matita -- recently also in Isabelle/HOL
    * but not in SMT-LIB (2.5?)

@@

\paragraph{Conventions.}

* monomorphic (or many-sorted) first-order logic
* Signature: $\Sig = (\Types, \Funcs)$

* types are simply atomic sorts, with no structure, and interpreted by nonempty
  domains
  * set of types must contain the distinguished \ty{bool} type of Booleans, interpreted
    as the set of truth values, and may contain other interpreted types (e.g., \ty{int}, \ty{real}).

* although nothing prevents composing the decision procedure with theories
  providing polymorphic types (parametric sorts), such as for arrays (e.g., $\ty{array}(\alpha,\beta)$)

* in keeping with SMT conventions, there are no separate predicate symbols;
  predicates are modeled as functions that return $\ty{Bool}$

* $\tau$ range over types---whether (co)datatypes or uninterpreted types
* $\delta$ range over (co)datatypes

* use sans-serif for function symbols (e.g., $\const{f}$) to distinguish them
from variables (e.g., $x$).

* function symbols are accompanied by a type signature, e.g. $\const{f} : \tau_1 \times \cdots \times \tau_n \to
\tau$ indicates takes $n \ge 0$ arguments of types $\tau_1,\ldots,\tau_n$ and returns a value of type $\tau$;
* applied as $\const{f}(t_1,\ldots,t_n)$ to well-typed arguments $t_1 : \tau_1$, \ldots,
$t_n : \tau_n$.

* nullary function symbols are called constants;
  following tradition, we simply write $\const{c} : \tau$ for the type signature 
  and $\const{c}$ rather than $\const{c}()$.

@@

We assume a signature $\Sig = (\Types, \Funcs)$.

Furthermore, a partition of the types:
  $\Types = \Data \uplus \Codata \uplus \Nondata$
with
  * $\Data$: datatypes
  * $\Codata$: codatatypes
  * $\Nondata$: non-(co)datatypes

And a partition of the functions:
  $\Funcs = \Ctr \uplus \Sel$
with
  * $\Ctr$: constructors
  * $\Sel$: selectors

(Do not need to consider other function symbols, because they would be
abstracted away as variables in the context of theory combination.)

In practice, e.g. in an SMT problem, we'd have:
  * declaration of uninterpreted types (in any order)
  * list of specifications of mutually (co)recursive groups of (co)datatypes

A (co)datatype specification consists of $l$~mutually recursive types which are
either all datatypes or all codatatypes. Each datatype $\delta$ is equipped by
$m \ge 1$ constructors, and each constructor for $\delta$ takes zero or more
arguments and returns a $\delta$ value. The argument types must be either
plain, be or be one of the newly introduced (co)datatypes. To every
argument corresponds a selector. The introduced names for the (co)datatypes, the
constructors, and the selectors must be distinct and different from already
introduced ones. Schematically:
%
\[
\begin{aligned}[t]
(\keyw{co})\keyw{datatype}\;\,
  \delta_1 & {}=\, \const{C}_{1}\rlap{${}_{1}$}\phantom{{}_{m}}(\bigl[\const{s}_{111}{:}\bigr]\; \tau_{111}, \,\ldots,\, \bigl[\const{s}_{11n_{11}}{:}\bigr]\; \tau_{11n_{11}}) & \\
           & {}\;\phantom{=\,}\;{} \smash{\,\,\,\vdots} \\[-\jot] % {}\;\phantom{=}\llap{\ensuremath{\mid}}\; \cdots \\
           & {}\;\phantom{=\,}\llap{\ensuremath{\,\mid\,}}\; \const{C}_{1m}(\bigl[\const{s}_{1m1}{:}\bigr]\; \tau_{1m1}, \,\ldots,\, \big[\const{s}_{1mn_{1m}}{:}\big]\; \tau_{1mn_{1m}}) \\[-1\jot]
   \smash{\vdots\,\,\,} \\[-1\jot]
  \keyw{and}\; \,\delta_l & =\, \ldots
\end{aligned}
\]
%
with
$\const{C}_{ij} : \tau_{ij1}\times\cdots\times\tau_{ijk_{ij}} \to \delta_i$
and $\const{s}_{ijk} : \delta_i \to \tau_{ijk}$. Defaults are assumed for
the selector names if they are omitted.

>>>>

\newcommand\Data{\Types_{\mathrm{dt}}}
\newcommand\Codata{\Types_{\mathrm{cdt}}}
\newcommand\Nondata{\Types_{\mathrm{oth}}}

@@

  * less briefly: codatatypes and why they are useful
    * in Agda, Coq, Matita -- recently also in Isabelle/HOL
    * but not in SMT-LIB 2 \cite{xxx}

@@

\paragraph{Conventions.}
This paper assumes a monomorphic (or many-sorted) first-order logic throughout.
A signature $\Sig = (\Types, \Funcs)$ consists of a set of types $\Types$ and a
set of function symbols $\Funcs$. Types are simply atomic sorts, with no
structure, and interpreted by nonempty domains. The set $\Types$ must contain a
distinguished type \ty{bool} interpreted as the set of truth
values. %, and may contain other interpreted types (e.g., \ty{int}, \ty{real}).
Predicates are simply functions that return $\ty{bool}$.
The metavariables $\delta, \tau$ range over types, with $\delta$ reserved
for (co)datatypes.

Function symbols are written in a sans-serif font (e.g., $\const{f}$, $\const{g}$) to
distinguish them from variables (e.g., $x$, $y$). With each function symbol \const{f}
is associated a list of argument types $\tau_1,\ldots,\tau_n$ (with $n \ge 0$)
and a return type $\tau$. This connection can be expressed compactly as
$\const{f} : \tau_1 \times \cdots \times \tau_n \to \tau$
(which collapses to $\const{f} : \tau$ if $n = 0$).
For a term $t$, the notation $t : \tau$ indicates that $t$ has type $\tau$.
Functions are applied the standard way, with $\const{f}(t_1,\ldots,t_n)$
applying \const{f} to $n$ well-typed arguments $t_1 : \tau_1$, \ldots, $t_n :
\tau_n$. Nullary function symbols are called constants and can appear as is in
terms, without parentheses.

%* although nothing prevents composing the decision procedure with theories
%  providing polymorphic types (parametric sorts), such as for arrays (e.g., $\ty{array}(\alpha,\beta)$)

@@

We assume a signature $\Sig = (\Types, \Funcs)$. The types are partioned into
$\Types = \Data \mathrel{\uplus} \Codata \mathrel{\uplus} \Nondata$, where $\Data$ are the
datatypes, $\Codata$ are the codatatypes, and $\Nondata$ are the remaining
types. The functions are partitioned into $\Funcs = \Ctr \mathrel{\uplus} \Sel$, where
$\Ctr$ are the constructors and $\Sel$ are the selectors. (There is no need to
consider other function symbols, because they can be abstracted away as
variables when combining theories.)

In practice (e.g., in an SMT problem), the signature is given by specifying the
uninterpreted types first (in any order), then the (co)datatypes with their
constructors and selectors in groups of mutually (co)recursive groups of
(co)datatypes, and finally any other function symbols.

A (co)datatype specification consists of $l$~mutually recursive types which are
either all datatypes or all codatatypes. Each datatype $\delta$ is equipped by
$m \ge 1$ constructors, and each constructor for $\delta$ takes zero or more
arguments and returns a $\delta$ value. The argument types must be either
plain, be or be one of the newly introduced (co)datatypes. To every
argument corresponds a selector. The introduced names for the (co)datatypes, the
constructors, and the selectors must be distinct and different from already
introduced ones.%
\footnote{In practice, it can be useful to specify the same selector
for several constructors of the same (co)datatype,
as long as the argument types coincide \cite[Section~3]{blanchette-et-al-2014-codata}.
However, this is not allowed by SMT-LIB 2.}
Schematically, using a syntax similar to Standard ML, Haskell,
or SMT-LIB:
%
\[
\begin{aligned}[t]
(\keyw{co})\keyw{datatype}\;\,
  \delta_1 & {}=\, \const{C}_{1}\rlap{${}_{1}$}\phantom{{}_{m}}(\bigl[\const{s}_{111}{:}\bigr]\; \tau_{111}, \,\ldots,\, \bigl[\const{s}_{11n_{11}}{:}\bigr]\; \tau_{11n_{11}}) & \\
           & {}\;\phantom{=\,}\;{} \smash{\,\,\,\vdots} \\[-\jot] % {}\;\phantom{=}\llap{\ensuremath{\mid}}\; \cdots \\
           & {}\;\phantom{=\,}\llap{\ensuremath{\,\mid\,}}\; \const{C}_{1m}(\bigl[\const{s}_{1m1}{:}\bigr]\; \tau_{1m1}, \,\ldots,\, \big[\const{s}_{1mn_{1m}}{:}\big]\; \tau_{1mn_{1m}}) \\[-1\jot]
   \smash{\vdots\,\,\,} \\[-1\jot]
  \keyw{and}\; \,\delta_l & =\, \ldots
\end{aligned}
\]
%
with
$\const{C}_{ij} : \tau_{ij1}\times\cdots\times\tau_{ijk_{ij}} \to \delta_i$
and $\const{s}_{ijk} : \delta_i \to \tau_{ijk}$. Defaults are assumed for
the selector names if they are omitted.
